A Refactoring - new attempt

### 2020-06-16 Articles/2020-Refactoring/Source/refactoring-article.tex
Starting to rewrite article on general theory

\documentclass[%
reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
amsmath,amssymb,
aps,
%pra,
prx,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-1}

%% Language and font encodings
%\usepackage[english]{babel}
%\usepackage[utf8x]{inputenc}
%\usepackage[T1]{fontenc}

%% Sets page size and margins
%\usepackage[letterpaper,top=1in,bottom=1in,left=1in,right=1in,marginparwidth=0.5in]{geometry}

%\usepackage{lineno,hyperref}
\usepackage{tikz}

% Theorem definitions using amsthm
\usepackage{amsmath, amsthm, amsfonts}

\newenvironment{rationale}{\emph{Rationale}.}{\qed}
\newenvironment{justification}{\emph{Justification}.}{\qed}
\renewenvironment{proof}{\emph{Proof}.}{\qed}

\theoremstyle{plain}% default 
\newtheorem{thrm}{Theorem}[section] 
\newtheorem{prop}[thrm]{Proposition} 
\newtheorem{coro}[thrm]{Corollary}

\theoremstyle{definition}
\newtheorem{defn}[thrm]{Definition}
\newtheorem{axiom}[thrm]{Axiom}
\newtheorem*{principle*}{Principle of scientific objectivity}
\theoremstyle{remark}
\newtheorem*{rem}{Remark}

\usepackage[only,llbracket, rrbracket,llparenthesis,rrparenthesis]{stmaryrd}

@@

% LOGIC symbols
% -------------

% Allows to create negation symbols
\let\vec\undefined
\usepackage{MnSymbol}

\DeclareMathOperator{\truth}{truth}
\DeclareMathOperator{\possFn}{poss}
\DeclareMathOperator{\result}{result}
\DeclareMathOperator{\idFn}{id}

\def\Bool{\mathbb{B}}
\def\TRUE{\textsc{true}}
\def\FALSE{\textsc{false}}

\def\SUCCESS{\textsc{success}}
\def\FAILURE{\textsc{failure}}
\def\UNDEF{\textsc{undefined}}

% Symbols for statements set
\def\stmtSet{\mathcal{S}}
\def\vstmtSet{\mathcal{S}_\textsf{v}}
\def\dstmtSet{\mathcal{S}_\textsf{d}}

% Symbols for tautology and contradiction
\def\tautology{\top}
\def\contradiction{\bot}

% Symbols for "compatibility" and "incompatibility"
\def\comp{\doublefrown}
\def\ncomp{\ndoublefrown}

% Symbols for "narrower" and "wider"
\def\narrower{\preccurlyeq}
\def\nnarrower{\npreccurlyeq}
\def\broader{\succcurlyeq}
\def\nbroader{\nsucccurlyeq}

% Symbol for "independent" and "correlated"
\def\indep{\upmodels}
\def\nindep{\nupmodels}

% Aliases for logical operations
\def\AND{\wedge}
\def\bigAND{\bigwedge}
\def\OR{\vee}
\def\bigOR{\bigvee}
\def\NOT{\neg}

% Formatting for statements
\newcommand{\stmt}[1][s] {\mathsf{#1}}
% Formatting for experimental tests
\newcommand{\expt}[1][e] {\mathsf{#1}}
\newcommand{\exptSet}{\mathcal{E}}
% Formatting for observations
\newcommand{\obs}[1][s] {\mathsf{#1}}

% Formatting for possibilities
\def\logCtx{\mathcal{S}}
\newcommand{\pAss}[1][\mathcal{S}] {\mathcal{A}_{#1}}

% Formatting for experimental domain
\newcommand{\edomain}[1][D] {\mathcal{#1}}

% Formatting for theoretical domain
\newcommand{\tdomain}[1][D] {\bar{\mathcal{#1}}}

\newcommand{\basis}[1][B] {\mathcal{#1}} % Basis

% Formatting for experimental relationships
\newcommand{\erel}[1][r] {#1}

% Formatting for sentence statements
\newcommand{\statement}[1] {\emph{``#1''}}

@@

\title{Towards a general mathematical theory of experimental science}

\author{Gabriele Carcassi}
\email{carcassi@umich.edu}
\author{Christine A. Aidala}
\affiliation{
	Physics Department \\
	University of Michigan \\
	Ann Arbor, MI 48109 \\
}

@@

\section{The principle of scientific objectivity}

As the starting point of our general theory, we introduce the following guiding principle:

\begin{principle*}
Science is universal, non-contradictory and evidence based.
\end{principle*}

This means that science concerns itself only to the study of assertions that have a well defined truth value which can be verified experimentally. The issue at hand is to formally capture this informal intuition. We begin with the following common definition.

\begin{defn}
	The \textbf{Boolean domain} is the set $\mathbb{B} = \{\FALSE, \TRUE\}$ of all possible truth values.
\end{defn}

Next we need to define our truth bearer. In mathematical logic, this is typically a well formed formula. This cannot work for us: what we are interested in is the meaning of the assertion and not how it is expressed. For example, \statement{this animal is a dog} and \statement{questo animale \`e un cane} represent the same fact expressed in different languages. As science is universal, it should not matter the language, units or reference system used to make an assertion.\footnote{In the same vein, a statement is not necessarily one sentence as it has to include all the information needed to understand what exactly we are talking about. These are the type of philosophical issues that are discussed at greater length in~\cite{Carc3}.} While we will still use standard mathematics for the formal system, we introduce a variation of algebraic logic to represent our ``informal'' assertions. We start by defining our truth bearer as:

\begin{axiom}\label{ax_statement}
	A \textbf{statement} $\stmt$ is an assertion that is either true or false. A \textbf{logical context} $\logCtx$ is a collection of statements with well defined logical relationships. Formally, a logical context $\logCtx$ is a collection of elements called statements upon which is defined a function $\truth: \logCtx \to \Bool$.
\end{axiom}

Note how the first part of the definition captures the informal meaning of what we are describing, while the second part captures the part that is formalized. This pattern will be present in most of our definitions and it serves to clarify both what is being formalized and how. Therefore, in science, the statement is an assertion while for the math it is just an element in some set.

While in math the truth value is the focus of the logic system, in science it is generally established experimentally. But our scientific model may constrain certain statements or statement combinations to be ruled out. For example, the statements \statement{this animal is a dog} and \statement{this animal is a cat} can both be either true or false, but they can't both be true. That is, the role of our logic system is not to keep track of what is true and false, but of what cannot possibly be true and cannot possibly be false. We need to keep track of these relationships so that we can never have inconsistencies or paradoxes.

\begin{axiom}\label{ax_possible_assignemetns}
	A \textbf{possible assignment} for a logical context $\logCtx$ is a map $a : \logCtx \to \Bool$ that assigns a truth value to each statement in a way consistent with the content of the statements. Formally, each logical context comes equipped with a set $\pAss \subseteq \Bool^{\logCtx}$ such that $\truth \in \pAss$. A map $a : \logCtx \to \Bool$ is a possible assignment for $\logCtx$ if $a \in \pAss$.
\end{axiom}

Note that the possible assignments are just hypotheticals. They are not actual physical entities. There is only one truth value, the one found experimentally. Logical consistency is defined on the truth assignments and not on the truth values. This allows us, for example, to track whether the meaning of each statement allows it to be true or not.

\begin{defn}
	A \textbf{tautology} $\tautology$ is a statement that is true in every possible assignment.
\end{defn}

\begin{defn}
	A \textbf{contradiction} $\contradiction$ is a statement that is false in every possible assignment.
\end{defn}

We also need to express relationships between the truth of different statements. For example, if we assign true to \statement{this animal is a dog} then we cannot assign true to \statement{this animal is not a dog}. Therefore we introduce the following:

\begin{axiom}\label{ax_functions_of_statement}
	We can always construct a statement whose truth value arbitrarily depends on an arbitrary set of statements. Formally, let $S \subseteq \logCtx$ be a set of statements and $f_{\Bool} : \Bool^S \to \Bool$ an arbitrary function from an assignment of $S$ to a truth value. Then we can always find a statement $\bar{\stmt} \in \logCtx$ such that
	$$a(\bar{\stmt}) = f_{\Bool}(\{a(\stmt)\}_{\stmt \in S})$$
	for every possible assignment $a \in \pAss$.
\end{axiom}

We will use the standard symbols $\NOT$, $\AND$, $\OR$ to indicate the negation (logical NOT), conjunction (logical AND) and disjunction (logical OR). With these three axioms, we can rederive all of the rules of classical logic. First we define our equivalence.

\begin{defn}
	Two statements $\stmt_1$ and $\stmt_2$ are \textbf{equivalent} $\stmt_1 \equiv \stmt_2$ if they must be equally true or false simply because of their content. Formally, $\stmt_1 \equiv \stmt_2$ if and only if $a(\stmt_1) = a(\stmt_2)$ for all $a \in \pAss$.
\end{defn}

From this definition one can prove the following two propositions.

\begin{prop}
	Statement equivalence satisfies the following properties:
	\begin{itemize}
		\item reflexivity: $\stmt \equiv \stmt$
		\item symmetry: if $\stmt_1 \equiv \stmt_2$ then $\stmt_2 \equiv \stmt_1$
		\item transitivity: if $\stmt_1 \equiv \stmt_2$ and $\stmt_2 \equiv \stmt_3$ then $\stmt_1 \equiv \stmt_3$
	\end{itemize}
	and is therefore an \textbf{equivalence relationship}.
\end{prop}

\begin{prop}\label{boolean_properties}
	The set of all statements $\mathcal{S}$ satisfies the following properties:
	\begin{itemize}
		\item associativity: $a \OR (b \OR c) \equiv (a \OR b) \OR c$, $a \AND (b \AND c) \equiv (a \AND b) \AND c$
		\item commutativity: $a \OR b \equiv b \OR a$, $a \AND b \equiv b \AND a$
		\item absorption: $a \OR (a \AND b) \equiv a$, $a \AND (a \OR b) \equiv a$
		\item identity: $a \OR \contradiction \equiv a
		$, $a \AND \tautology \equiv a$
		\item distributivity: $a \OR (b \AND c) \equiv (a \OR b) \AND (a \OR c)$, $a \AND (b \OR c) \equiv (a \AND b) \OR (a \AND c)$
		\item complements: $a \OR \NOT a \equiv \tautology$, $a \AND \NOT a \equiv \contradiction$
		\item De Morgan: $\NOT a \OR \NOT b \equiv \NOT (a \AND b)$, $\NOT a \AND \NOT b \equiv \NOT (a \OR b)$
	\end{itemize}
	This, by definition, means $\mathcal{S}$ is a \textbf{Boolean algebra}.
\end{prop}

One can also prove that the Boolean algebra is complete, and therefore it generalizes to the infinite case.

This system not only allows us to use formal classical logic while keeping the statements informal, but is also equipped to capture causal relationships.\footnote{Note that the axioms prevent any form of paradox, simply because a possible assignment must exist. Therefore situations like \statement{this statement is not true} are ruled out simply because they cannot satisfy the axioms.} For example, $\stmt_1=$\statement{the thermometer indicator is between 24 C and 25 C} and $\stmt_2=$\statement{the temperature is between 24 C and 25 C} are equivalent because one is true if and only if the other is true (on the assumption that our thermometer is actually working). We can also define other semantic relationships.

\begin{defn}
	Given two statements $\stmt_1$ and $\stmt_2$, we say that:
	\begin{itemize}
		\item $\stmt_1$ \textbf{is narrower than} $\stmt_2$ (noted $\stmt_1 \narrower \stmt_2$) if $\stmt_2$ is true whenever $\stmt_1$ is true.
		\item $\stmt_1$ \textbf{is broader than} $\stmt_2$ (noted $\stmt_1 \broader \stmt_2$) if $\stmt_2 \narrower \stmt_1$.
		\item $\stmt_1$ \textbf{is compatible to} $\stmt_2$ (noted $\stmt_1 \comp \stmt_2$) if there is an assignment in which they are both true.

	\end{itemize}
	The negation of these properties will be noted by $\nnarrower$, $\nbroader$ , $\ncomp$ respectively.
\end{defn}
\begin{defn}
	The elements of a set of statements $S \subseteq \logCtx$ are said to be \textbf{independent} (noted $\stmt_1 \indep \stmt_2$ for a set of two) if the assignment of any subset of statements does not depend on the assignment of the others. That is, a set of statements $S \subseteq \logCtx$ is independent if given a family  $\{t_{\stmt}\}_{\stmt \in S}$ such that each $t_{\stmt} \in \Bool$ is a possible assignment for the respective $\stmt$ we can always find $a \in \pAss$ such that $a(\stmt) = t_{\stmt}$ for all $\stmt \in S$.
\end{defn}

For example, given the following statements:
\begin{description}
	\item $\stmt_1=$\statement{that animal is a cat}
	\item $\stmt_2=$\statement{that animal is a mammal}
	\item $\stmt_3=$\statement{that animal is a dog}
	\item $\stmt_4=$\statement{that animal is black}
\end{description}
we have the $\stmt_1$ $\narrower$ $\stmt_2$ (i.e. it is more specific), $\stmt_1$ $\ncomp$ $\stmt_3$ (i.e. they cannot be true at the same time) and $\stmt_1$ $\indep$ $\stmt_4$ (i.e. the truth of one tells us nothing about the truth of the other).

Another interesting result is that statement narrowness imposes a partial order on the set of all statements.

\begin{prop}
	Statement narrowness satisfies the following properties:
	\begin{itemize}
		\item reflexivity: $s \narrower s$
		\item antisymmetry: if $s_1 \narrower s_2$ and  $s_2 \narrower s_1$ then $s_1 \equiv s_2$
		\item transitivity: if $s_1 \narrower s_2$ and $s_2 \narrower s_3$ then $s_1 \narrower s_3$
	\end{itemize}
	and is therefore a \textbf{partial order}.
\end{prop}

As every element in our general theory will be constructed upon statements, these operations are important as they will characterize all those constructions. For example, if we quantify the precision of a set of statements, it will need to be ordered in a way that is compatible with narrowness. If we define statistical independence, it will have to be defined in a way that is compatible with statement independence. In the same way that set theory defines concepts common across all mathematics, the general theory defines basic concepts that are common to all scientific theories.

\section{Verifiable statements and experimental domains}

We have the tools for dealing with assertions that are universal and non-contradictory, now we have to develop the tools for those that are evidence based as well.

\begin{axiom}\label{ax_verifiable_statements}
	A \textbf{verifiable statement} is a statement that can be shown to be true experimentally. Formally, a statement $\stmt$ is verifiable if it is part of the subset $\stmt \in \vstmtSet \subset \stmtSet$ of all verifiable statements.
\end{axiom}

Physically, this means that we have a repeatable procedure that anybody can execute and that always gives the same result. If the statement is true, then this experimental test must terminate successfully. Note that, in general, the test may not terminate if the statement is false. Consider the following:
\begin{enumerate}
	\item find a swan
	\item if it is black terminate successfully
	\item go to step 1
\end{enumerate}
This will terminate if a black swan is found, but it will not terminate if no black swans exist (i.e. absence of evidence is not evidence of absence). Because of non-termination, failure to verify is not verification of the negation. That is, the negation of a verifiable statement is not necessarily a verifiable statement.

Given two statements, though, we can verify their conjunction simply by verifying both: if they are both true, both their tests will terminate and verify the conjunction. But we cannot extend this to an infinite number of statements as we would never terminate. We can also verify the disjunction of two statements: once one test terminates we are done. And because we only need one test to terminate, we can generalize to a countable number of verifiable statements by following this procedure:
\begin{enumerate}
	\item initialize $n$ to 1
	\item for each $i=1..n$
	\begin{enumerate}
		\item run the test for $\stmt_i$ for $n$ seconds
		\item if it terminates successfully then terminate successfully
	\end{enumerate}
	\item increment $n$ and go to step 2
\end{enumerate}
This procedure will run all tests for an arbitrary amount of time. Therefore, if one statement is true, it will make the test terminate successfully, even if all other tests would never terminate. In light of this, we can set the following axioms.

\begin{axiom}\label{ax_verifiable_AND}
	The conjunction of a finite collection of verifiable statements is a verifiable statement. Formally, let $\{\stmt_i\}_{i=1}^{n} \subseteq \vstmtSet$ be a finite collection of verifiable statements. Then the conjunction $\bigAND\limits_{i=1}^{n} \stmt_i \in \vstmtSet$ is a verifiable statement.
\end{axiom}
	\begin{axiom}\label{ax_verifiable_OR}
	The disjunction of a countable collection of verifiable statements is a verifiable statement. Formally, let $\{\stmt_i\}_{i=1}^{\infty} \subseteq \vstmtSet$ be a countable collection of verifiable statements. Then the disjunction $\bigOR\limits_{i=1}^{\infty} \stmt_i \in \vstmtSet$ is a verifiable statement.
\end{axiom}

We could also define decidable statements as those for which falsehood can also be tested experimentally. Table \ref{tab:algebras} compares the different algebras.

\begin{table*}
	\centering
	\begin{tabular}{p{0.14\textwidth} p{0.08\textwidth} p{0.13\textwidth} p{0.22\textwidth} p{0.23\textwidth}}
		Operator & Gate & Statement & Verifiable Statement & Decidable Statement  \\ 
		\hline 
		Negation & NOT & allowed & disallowed & allowed \\ 
		Conjunction & AND & arbitrary  & finite & finite \\ 
		Disjunction & OR & arbitrary  & countable & finite \\ 
	\end{tabular}
	\caption{Comparing algebras of statements.}\label{tab:algebras}
\end{table*}

Now that we can verify statements one by one, we need to define what it means to verify a group of them. In principle, given a set of verifiable statements we can simply start testing them one after the other. However, if we are given a set of uncountable statements then, even if we have an indefinitely long time at our disposal, we will not be able to create a procedure that eventually reaches all statements. However, we may not need to actually run all tests for all statements. For example, if we found that $\stmt_1$ is true then there is no need to test any of its disjunctions like $\stmt_1 \OR \stmt_2$.

\begin{defn}
	Given a set $\edomain$ of verifiable statements, $\basis \subseteq \edomain$ is a \textbf{basis} if the truth values of $\basis$ are enough to deduce the truth values of the set. Formally, all elements of $\edomain$ can be generated from $\basis$ using finite conjunction and countable disjunction.
\end{defn}

Therefore it is the size of the basis that matters and not the size of the set. If the basis is countable we can keep going and, if any statement is true, it will eventually be verified experimentally.\footnote{We assume we are given an indefinitely long time because this handles limits and because we have not given constraints for how long or short a successful test can be (i.e. if we ran a test for three days without terminating, it may still successfully complete in three days and one minute).}

\begin{defn}
	An \textbf{experimental domain} $\edomain$ represents all the experimental evidence that can be acquired about a scientific subject in an indefinite amount of time. Formally, it is a set of statements, closed under finite conjunction and countable disjunction, that includes precisely the tautology, the contradiction, and a set of verifiable statements that can be generated from a countable basis.
\end{defn}

These axioms and definitions formally characterize what we mean by evidence based. A scientific theory will be fully defined by a countable set of verifiable statements, the basis for an experimental domain.

\section{Theoretical domains and possibilities}

While verifiable statements define what can be scientifically studied, not all interesting scientific statements are directly verifiable. Consider the two statements \statement{there exists extra-terrestrial life} and \statement{there is no extra-terrestrial life}. We can verify the first if we happen to find signs of life somewhere, but experimentally verifying the second is practically impossible. Yet, the second is still meaningful as a prediction: it predicts that the test for the first will never terminate. That is, while negations are not verifiable we can still logically talk about them when, for example, constructing truth assignments.

\begin{defn}
	The \textbf{theoretical domain} $\tdomain$ of an experimental domain $\edomain$ is the set of statements that we can use to state predictions, which is constructed from $\edomain$ by allowing negation. We call \textbf{theoretical statement} a statement that is part of a theoretical domain. More formally, $\tdomain$ is the set of all statements generated from $\edomain$ using negation, finite conjunction and countable disjunction.
\end{defn}

@@

\begin{defn}
	Let $\edomain$ be an experimental domain and $X$ its possibilities. We define the map $U : \edomain \rightarrow 2^X$ that for each statement $\obs \in \edomain$ returns the set of possibilities compatible with it. That is, $U(\obs)\equiv\{ x \in X \, | \, x \comp \obs\}$. We call $U(\obs)$ the \textbf{verifiable set} of possibilities associated with $\obs$.
\end{defn}

>>>>

\documentclass[letterpaper]{article}

\input{../../../include/basicstyle}
\input{../../../include/theorems}
\input{../../../include/logic}

\usepackage{tikz}

@@

@@

\title{A Refactoring - new attempt}

\author{Gabriele Carcassi}
%\email{carcassi@umich.edu}
\author{Christine A. Aidala}
%\affiliation{
%	Physics Department \\
%	University of Michigan \\
%	Ann Arbor, MI 48109 \\
%}

@@

\section{A standard for precision?}

Mathematics has evolved a set of rules and standards for what constitutes a well-posed mathematical theory. There is no such standard in science in general and in physics in particular. In fact, given that some fields are mostly driven by data and heuristics while others are driven by mathematical formalism, we have found divergent opinions in what can and should be formalized. We therefore had to develop our own standard, which is based upon the following insights.

\textit{Formal systems cannot fully capture scientific theories.} The same mathematical structure (e.g. the equation $a= bc$) can refer to multiple physical relationships (e.g. Newton's second law or Ohm's law) which cannot be reconstructed solely on the basis of the formal relationships. Moreover, qualitative reasoning in many fields can be carried out independently. In this light, formal systems are a poor starting point for a scientific theory.

\textit{Formal systems in science should be justified, not interpreted.} We want to start from the informal definitions of the physical objects, be clear as to what aspects are formalized and why that formalization follows. The formal system, then, is a tool useful to check for consistency and clarification of particular aspects of the theory. It is useful only insofar its connection to the physical content is clear.

Our framework will therefore consist of both informal and formal parts. We distinguish two ways new concepts are introduced: axioms and definitions. \textbf{Axioms} introduce new elements: they define new concepts and what aspects are captured in the formal system.  \textbf{Definitions} further characterize already introduced objects, without introducing new elements in the formal system. Both axioms and definitions will consist of two parts. The first gives the intuitive, informal description while the second gives the formal characterization. The formal part will necessarily not capture the entirety of the informal concept. Where needed, a \textbf{justification} will accompany axioms and definitions. These are informal arguments that explain why the formal objects must have those formal properties. All \textbf{propositions} and \textbf{theorems} are declared and proven in the formal system, and they deal with objects that are now both mathematically and scientifically well specified.

The overall goal is to keep the axioms to a minimum and to a level of simplicity that is easily checked. This minimizes the part of the framework that can potentially be open to inconsistencies.

Axioms, definitions and justifications are included in the mathematical section, with the rest of the formal elements. They should be seen as integral part of the framework.

\section{Basic formal structures?}

The starting point of our framework is the \textbf{principle of scientific objectivity} which states \statement{Science is universal, non-contradictory and evidence based}. Every physical theory, then, must give us a set of logically sound statements linked with some notion of experimental verification. The content of the statements, what can be verified and how, will vary from theory to theory, but logical consistency and verifiability must be common to all of them. Therefore we will have two set of axioms: the \textbf{logical consistency axioms} and the \textbf{experimental verifiability axioms}. We start with the first set.

\subsection{Logical consistency}

The \textbf{axiom of context} establishes that \textbf{statements} are organized into \textbf{logical contexts}. It is the context that defines their meaning and logical relationships. The definition of a physical object, for example, may change depending on our goal: if we measure the mass of the electron, its value is assumed to be unknown; if we perform particle identification in a detector, the mass of the electron is assumed to be a particular value. Each context has to be such that the meaning of each statement has no ambiguity and therefore it is associated with a single, well-defined truth value. This is, in general, not known: it cannot be proven or deduced. Ultimately, it is the role of experimentation to find it.

The \textbf{axiom of possibility} establishes that not all truth assignments are consistent with the meaning of the statements. For example, it is not possible for \statement{the particle is an electron} and \statement{the particle is a positron} to be true at the same time. As the context must fully specify the meaning of each statement, all these relationships must be well-specified. A context therefore will define the \textbf{possible assignments} for all statements, and the actual truth must be one of them.

Lastly, the \textbf{axiom of closure} establishes that we can always find a statement whose truth depends on the truth of others through a chosen Boolean function. For example, we can find the negation of a statement, the conjunction (logical AND) of any two statements, and so on.

These three axioms introduce the basis of a logic system all scientific theories must adhere to. We can visualize a logical context as a table. Suppose we have a context $\logCtx$ which contains at least the following statements:
\begin{description}
	\item $\stmt_1=$\statement{that animal is an animal}
	\item $\stmt_2=$\statement{that animal is a cat}
	\item $\stmt_3=$\statement{that animal is a mammal}
	\item $\stmt_4=$\statement{that animal has three bones in the middle ear}
	\item $\stmt_5=$\statement{that animal is a dog}
	\item $\stmt_6=$\statement{that animal is black}
	\item $\stmt_7=$\statement{that animal is a plant}
\end{description}
We would have:
\begin{center}
	\begin{tabular}{c|c|c|c|c|c|c|c}
		$\stmt_1$ & $\stmt_2$ & $\stmt_3$ & $\stmt_4$ & $\stmt_5$ & $\stmt_6$ & $\stmt_7$ & ... \\
		\hline
		T & T & T & T & F & T & F & ... \\
		T & T & T & T & F & F & F & ... \\
		T & F & T & T & F & T & F & ... \\
		T & F & F & F & T & T & F & ... \\
		T & F & F & F & F & F & F & ... \\
		... & ... & ... & ... & ... & ... & ... \\
	\end{tabular}
\end{center}
where each line represents a possible assignment, each column a statement and each cell the truth value for each statement under the given assignment.

We can give some further definitions, which we introduce with examples based on the previous table. We say $\stmt_3$ and $\stmt_4$ are \textbf{equivalent}, noted $\stmt_3 \equiv \stmt_4$ because they have the same truth value in each assignment (only mammals have three bones in the middle ear). We say $\stmt_2$ is \textbf{narrower} than $\stmt_3$ (i.e. more specific), noted $\stmt_2 \narrower \stmt_3$, because whether the first one is true the second one is true as well (if the animal is a cat, it is also a mammal). We say $\stmt_2$ and $\stmt_5$ are not \textbf{compatible}, noted $\stmt_2 \ncomp \stmt_5$, because there is no assignment in which they are both true (if the animal is a cat, it is not a dog). We say $\stmt_2$ is independent from $\stmt_6$, noted $\stmt_2 \indep \stmt_6$, because the truth value of one does not constrain the truth value of the other (knowing whether the animal is a cat or not does not tell us whether it is black or not). We say $\stmt_1$ is a \textbf{tautology} $\tautology$, noted $\stmt_1 \equiv \tautology$, because it is true in every assignment. We say $\stmt_7$ is a \textbf{contradiction} $\contradiction$, noted $\stmt_7 \equiv \contradiction$, because it is false in every assignment. All others are \textbf{contingent}.

From now on, we will not differentiate between two statements that are equivalent. That is, we will implicitly refer to the equivalence classes. With that in mind, one can show that a logical context is a \textbf{complete Boolean algebra} and that $\narrower$ is a partial order, the one corresponding to the lattice of the algebra. While this should not be surprising, the point is that we start from premises that are easily justified and that give a clear understanding of what the formal object represents.

This will be essential when adding new definitions. Any other formal structure imposed on the statements, in fact, is capturing other properties of the same objects and will have to ``play nice'' with the above logical structure. This means that the notion of equivalence will, in practice, become an isomorphism within the new structure; the ordering imposed by numeric quantities will be a specialized version of the one imposed by $\narrower$; the notion of independence will relate to the notion of independence in statistic, probability, linear algebra, dynamical system, and so on. Understanding this simple logical structure, then, means getting insights in many different areas and in how they are ultimately connected.

% In mathematical logic, this is typically a well formed formula. This cannot work for us: what we are interested in is the meaning of the assertion and not how it is expressed. For example, \statement{this animal is a dog} and \statement{questo animale \`e un cane} represent the same fact expressed in different languages. As science is universal, it should not matter the language, units or reference system used to make an assertion.\footnote{In the same vein, a statement is not necessarily one sentence as it has to include all the information needed to understand what exactly we are talking about. These are the type of philosophical issues that are discussed at greater length in~\cite{Carc3}.} While we will still use standard mathematics for the formal system, we introduce a variation of algebraic logic to represent our ``informal'' assertions. We start by defining our truth bearer as:

%Note how the first part of the definition captures the informal meaning of what we are describing, while the second part captures the part that is formalized. This pattern will be present in most of our definitions and it serves to clarify both what is being formalized and how. Therefore, in science, the statement is an assertion while for the math it is just an element in some set.

%Note that the possible assignments are just hypotheticals. They are not actual physical entities. There is only one truth value, the one found experimentally. Logical consistency is defined on the truth assignments and not on the truth values. This allows us, for example, to track whether the meaning of each statement allows it to be true or not.

%This system not only allows us to use formal classical logic while keeping the statements informal, but is also equipped to capture causal relationships.\footnote{Note that the axioms prevent any form of paradox, simply because a possible assignment must exist. Therefore situations like \statement{this statement is not true} are ruled out simply because they cannot satisfy the axioms.} For example, $\stmt_1=$\statement{the thermometer indicator is between 24 C and 25 C} and $\stmt_2=$\statement{the temperature is between 24 C and 25 C} are equivalent because one is true if and only if the other is true (on the assumption that our thermometer is actually working). We can also define other semantic relationships.

\subsection{Experimental verifiability}

Having seen the logical consistency axioms, we turn our attention to the experimental verifiability axioms.

The \textbf{axiom of verifiability} establishes that some statement are experimentally \textbf{verifiable}, meaning we have a test at our disposal that, if the statement is true, will terminate successfully in finite time. If the statement is not true, the test may terminate unsuccessfully or never terminate. To give an example, consider the following:
\begin{enumerate}
	\item find a swan
	\item if it is black terminate successfully
	\item go to step 1
\end{enumerate}
This will terminate if a black swan is found, but it will not terminate if no black swans exist (i.e. absence of evidence is not evidence of absence). The axiom also states that tautologies and contradictions are verifiables (i.e. they can be associated with trivial tests) and that if one statement is equivalent to a verifiable one, then it is also verifiable (i.e. we can use the test of the second since either both statements are true or both are false).

Termination in finite time is important as it imposes a different logical structure.\footnote{This is broadly recognized in computer science theory and in constructive logic.} Because of non-termination, failure to verify is not verification of the negation. That is, the negation of a verifiable statement is not necessarily a verifiable statement.

The \textbf{axiom of finite conjunction verifiability} establishes that, given a finite set of statements, we can verify their conjunction (i.e. the logical AND). In fact, we can execute one test at a time, and if they are all successful, they will be so in finite time therefore the conjunction will be verified as well. This will not extend to the infinite case.

The \textbf{axiom of countable disjunction verifiability} establishes that, given a countable set of statements, we can verify their disjunction (i.e. the logical OR). In fact, if the disjunction is true then one statement must be true. Once that is verified, we can terminate regardless of how many other statements are there in the disjunction. The set of statements must be countable, though, or it may take infinite time to find the statement that is true.

On top of these axioms we can define \textbf{decidable statements} as the ones that can be both verified and falsified experimentally (i.e. their negation is a verifiable statement). One can prove that the finite disjunction and the finite conjunction of decidable statements is also decidable.

We therefore have three types of algebras embedded one into the other. The logical context $\logCtx$ is a complete Boolean algebra; the set of all verifiable statement $\vstmtSet \subset \logCtx$ is, in general, not a Boolean algebra but a complete Heyting algebra; the set of all decidable statements $\dstmtSet \subset \vstmtSet \subset \logCtx$ is a (finite) Boolean algebra. The interplay between these structures is important and leads to topologies and $\sigma$-algebras, the foundations for geometry, Lie algebras, measure theory, probability theory and most of the mathematical structures widely used in physics.

\subsection{Domains and possibilities?}

Finite time termination not only imposes a different logical structure, but it also limits the number of statements that can be verified. Even when given an arbitrarily long time, a countable set of statements is the most that can be verified. However, due to the logical relationship between statements, one does not need to test all of them: if \statement{the mass of the photon is less than $10^{-6}$ eV} is experimentally verified, there is no need to test whether \statement{the mass of the photon is greater than $10^{-6}$ eV}.

Given a set $\edomain$ of verifiable statements, we call a \textbf{basis} $\basis \subset \edomain$ a subset whose tests are enough to verify the whole set. Formally, $\edomain$ can be generated from $\basis$ using finite conjunction and countable disjunction. We say $\edomain$ is an \textbf{experimental domain} if it is a maximal set (i.e. includes all the verifiable statements) generated from a countable basis.

An experimental domain represents all experimental evidence that can be acquired in an indefinite amount of time. It represents the most fundamental object in our framework and, if our scientific theory is evidence based, it must be fully specified by an experimental domain.

Not all statements in a scientific theory, though, are verifiable statement. For example, \statement{there is no extra-terrestrial life} is not something we can verified (at best we can only conclude we have found none where we looked), and neither is \statement{the mass of the photon is exactly 0 eV} (at best we can exclude is above a certain threshold). In general, while negations are not necessarily verifiable, we can still logically talk about them and they must be included in our theory.

Given an experimental domain $\edomain$ we define its \textbf{theoretical domain} $\tdomain$ which includes all statements generated by $\edomain$ allowing negations as well. Theoretical domains are $\sigma$-complete Boolean algebras as they are closed under negation and countable conjunction and disjunction.

For each theoretical statement $\tstmt \in \tdomain$, we can define its \textbf{verifiable part} $\ver(\tstmt)$, \textbf{falsifiable part} $\fal(\tstmt)$ and its \textbf{undecidable part} $\und(\tstmt)$. This allows us to precisely characterize the optimal testing strategy in terms of termination. For example, for a decidable statement like $\stmt=$\statement{there are three apples on the table}, the verifiable part will be equal to the statement itself $\ver(\stmt) \equiv \stmt$, the falsifiable part will be equal to its negation cases $\fal(\stmt) \equiv \NOT\stmt$ and the undecidable part will be the contradiction $\und(\stmt) \equiv \contradiction$. The experimental test for $\stmt$ always terminates either successfully or unsuccessfully. Instead, if $\stmt=$\statement{the mass of the photon is less than $10^{-17}$ eV}, we still have $\ver(\stmt) \equiv \stmt$ because it is a verifiable statement, but we will have $\fal(\stmt) \equiv$\statement{the mass of the photon is more than $10^{-17}$ eV} while $\und(\stmt) \equiv$\statement{the mass of the photon is exactly $10^{-17}$ eV}. The issue is that, due to finite precision, if the mass of the photon where exactly $10^{-17}$ eV, our confidence interval would only include the region below that number, so we would never be able to exclude it. So, the test would not terminate in that particular case. Lastly, if $\stmt=$\statement{the mass of the photon is a rational number as expressed in eV}, we have $\ver(\stmt) \equiv \contradiction \equiv \fal(\stmt)$ and $\und(\stmt) \equiv \tautology$. The only way to know whether the statement is true or not would be to know the exact value of the mass, which cannot be achieved in finite time. Therefore the test for $\stmt$ will never terminate. We call such statements \textbf{undecidable}.

The fact that a scientific theory can contain undecidable statements, statements about which nothing can be said experimentally, may seem counter-intuitive. Theoretical statements include all those statement for which, in principle, a test can be devised, regardless of whether the test actually terminates: the $\sigma$-completeness guarantees that. Note that if we closed under arbitrary union, not just countable, we would bring in statements for which a test cannot even be, in principle, constructed. These would be unphysical objects, and are excluded by our framework.\footnote{These would correspond to non-Borel sets.}

@@

\begin{defn}
	Let $\edomain$ be an experimental domain and $X$ its possibilities. We define the map $U : \edomain \rightarrow 2^X$ that for each statement $\stmt \in \edomain$ returns the set of possibilities compatible with it. That is, $U(\stmt)\equiv\{ x \in X \, | \, x \comp \stmt\}$. We call $U(\stmt)$ the \textbf{verifiable set} of possibilities associated with $\stmt$.
\end{defn}
